### 2020-07-08 submission/manuscript.tex
Text edits before reorganization

\section{\texorpdfstring{Initial gut microbiota and response to
antibiotic perturbation influence \emph{Clostridioides difficile}
colonization in
mice}{Initial gut microbiota and response to antibiotic perturbation influence Clostridioides difficile colonization in mice}}\label{initial-gut-microbiota-and-response-to-antibiotic-perturbation-influence-clostridioides-difficile-colonization-in-mice}

@@

The microbiota plays a key role in determining susceptibility to
\emph{Clostridioides difficile} infections (CDIs). However, much of the
mechanistic work examining CDIs in mouse models use a single university
colony or vendor. We treated mice from 6 different colony sources (2
University of Michigan colonies and 4 vendors) with a single clindamycin
dose, followed by \emph{C. difficile} challenge 1 day later and measured
\emph{C. difficile} colonization levels through 9 days post-infection.
The microbiota was profiled via 16S rRNA gene sequencing analysis to
examine variation across colony sources and alterations due to
clindamycin treatment and \emph{C. difficile} challenge. While all
sources of mice were colonized 1-day post-infection, variation in
\emph{C. difficile} colonization levels emerged from days 3-7
post-infection with some sources colonized with \emph{C. difficile} for
slightly longer and at higher levels. We identified bacterial taxa with
different relative abundances across colony sources throughout the
experiment, as well as taxa that were consistently impacted by
clindamycin treatment in all sources of mice. We created logistic
regression models that successfully classified mice based on whether
they cleared \emph{C. difficile} by 7 days post-infection using
baseline, post-clindamycin, and post-infection community composition
data. After examining the taxa that were most important to the
classification models, we identified a subset of key taxa that varied
across colony sources (\emph{Bacteroides}, \emph{Deferribacteraceae}),
were altered by clindamycin (\emph{Porphyromonadaceae},
\emph{Ruminococcaceae}), or both (\emph{Enterobacteriaceae},
\emph{Enterococcus}, \emph{Bifidobacteriaceae},
\emph{Coriobacteriaceae}, \emph{Lachnospiraceae}, and
\emph{Verrucomicrobiaceae}). These results suggest the response of the
initial gut microbiota to clindamycin treatment influences \emph{C.
difficile} 630 colonization dynamics.

@@

\emph{Clostridioides difficile} is a leading nosocomial infection.
Although the microbiota has been established as a key risk factor, there
is variation in who becomes asymptomatically colonized, develops an
infection, or has an infection with adverse outcomes. \emph{C.
difficile} infection (CDI) mouse models are widely used to answer a
variety of \emph{C. difficile} pathogenesis questions. However, the
inter-individual variation between mice is less than what is observed in
humans, particularly if just one source of mice is used. In this study,
we administered clindamycin to mice from 6 different colony sources and
challenged them with \emph{C. difficile}. Interestingly, only a subset
of the taxa that vary across sources were associated with how long
\emph{C. difficile} was able to colonize. Future studies examining the
interplay between the microbiota and \emph{C. difficile} should consider
using mice from multiple sources to narrow down the microbes driving the
observed phenotypes and reflect human interindividual variation.

@@

Antibiotics are a clear risk factor for \emph{Clostridioides difficile}
infections (CDIs), but there is variation in who goes on to develop
severe or recurrent CDIs after exposure (1, 2). Additionally,
asymptomatic colonization, where \emph{C. difficile} is detectable, but
symptoms are absent has been documented in infants and adults (3, 4).
The intestinal microbiome has been implicated in asymptomatic
colonization (5, 6), susceptibility to CDIs (7), and adverse CDI
outcomes (9--12).

Mouse models of CDIs have been a great tool for understanding \emph{C.
difficile} pathogenesis (13). The number of CDI mouse model studies has
grown substantially since Chen et al. published their C57BL/6 model in
2008, which disrupted the gut microbiota with antibiotics to enable
\emph{C. difficile} colonization and symptoms such as diarrhea and
weight loss (14). CDI mouse models have been used to examine
translationally relevant questions regarding \emph{C. difficile},
including the role of the microbiota and efficacy of potential
therapeutics for treating CDIs (15). However, microbiome variation
between lab mice is much less than the variation observed between humans
(16, 17). Additionally, studying the contribution of the microbiota to a
particular disease phenotype in one set of lab mice after the same
perturbation could yield a number of findings of which only a fraction
may be driving the phenotype.

In the past, our group has attempted to introduce more microbiome
variation into the CDI mouse model by using a variety of antibiotic
treatments (18--21). An alternative approach to maximize microbiome
variation is to use mice from multiple sources (22, 23). Microbiome
differences between different mouse vendors have been well documented
and shown to influence susceptibility to a variety of diseases (24, 25),
including enteric infections (22, 23, 26--30). Additionally, different
research groups have observed different CDI outcomes in mice despite
using similar models and the microbiome has been proposed as one factor
potentially mediating CDI susceptibility and outcomes (13, 18, 21,
31--33). Here we examined how variations in the baseline microbiome and
responses to clindamycin treatment in C57BL/6 mice from six different
sources influenced susceptibility to \emph{C. difficile} colonization
and the time needed to clear the infection.

@@

\textbf{Clindamycin treatment renders all mice susceptible to \emph{C.
difficile} 630 colonization regardless of colony source.} To test how
the microbiotas of mice from different colony sources impact
colonization dynamics after clindamycin exposure, we utilized C57BL/6
mice from 6 different sources: two colonies from the University of
Michigan (the Young and Schloss lab colonies), the Jackson Laboratory,
Charles River Laboratories, Taconic Biosciences, and Envigo (which was
formerly Harlan). These 4 vendors were chosen because they represent
commonly used vendors for CDI studies in mice (26, 34--40). After a
13-day acclimation period for the mice ordered from vendors, all mice
were treated with 10 mg/kg clindamycin via intraperitoneal injection and
one day later challenged with 10\textsuperscript{3} \emph{C. difficile}
630 spores (Fig. 1A). Clindamycin was chosen because we have previously
demonstrated mice are rendered susceptible, but consistently cleared the
CDI within 9 days (21, 41), clindamycin is frequently implicated with
human CDIs (42), and is also part of the antibiotic treatment for the
frequently cited 2008 CDI mouse model (14). The day after infection,
\emph{C. difficile} was detectable in all mice at a similar level
(median CFU range: 2.2e+07-1.3e+08; \emph{P}\textsubscript{FDR} = 0.15),
indicating clindamycin rendered all mice susceptible regardless of
colony source (Fig. 1B). Interestingly, variation in \emph{C. difficile}
CFU levels across sources of mice emerged from days 3-7 post-infection
(all \emph{P}\textsubscript{FDR} \(\le\) 0.019; Fig. 1B and Table S1),
suggesting mouse colony source is associated with \emph{C. difficile}
clearance. We conducted two experiments approximately 3 months apart and
while the colonization dynamics were similar across most sources of
mice, there was some variation between the 2 experiments, particularly
for the Schloss and Envigo mice (Fig. S1A-B). Although \emph{C.
difficile} 630 causes mild symptoms in mice comparied to other \emph{C.
difficile} strains (43), we also saw that weight change significantly
varied across sources of mice with the most weight lost two days
post-infection (Fig. 1C and Table S2). Importantly, there was also one
Jackson and one Envigo mouse that died between 1- and 3-days
post-infection during the second experiment. Interestingly, mice ordered
from Jackson, Taconic, and Envigo tended to lose more weight (although
there was variation between experiments with Schloss and Envigo mice),
have higher \emph{C. difficile} CFU levels and take longer to clear the
infection compared to the other sources of mice, which was particularly
evident 7 days post-infection (Fig. 1B-C, Fig. S2C-D), when 57\% of the
mice were still colonized with \emph{C. difficile} (Fig. S1E). By 9 days
post-infection the majority of the mice from all sources had cleared
\emph{C. difficile} (Fig. S1C) with the exception of 1 Taconic mouse
from the first experiment and 2 Envigo mice from the second experiment.
Thus, clindamycin rendered all mice susceptible to \emph{C. difficile}
630 colonization, regardless of colony source, but variation across
sources emerged with 3 out of 6 sources taking longer to clear \emph{C.
difficile}.

\textbf{Bacterial communities consistently vary across mouse colony
sources despite antibiotic and infection perturbations.} Given the well
known variation in mouse microbiomes across vendors and university
colonies (25), we hypothesized that the variation in \emph{C. difficile}
clearance could be explained by microbiota variation across the 6
sources. We used 16S rRNA gene sequencing to characterize the fecal
bacterial communities from the mice over the course of the experiment.
Since antibiotics and other risk factors of CDIs are associated with
decreased microbiota diversity (44), we first examined alpha diversity
measures across the 6 sources of mice. Examining the bacterial
communities at baseline, prior to clindamycin treatment there was a
significant difference in the number of observed OTUs
(\emph{P}\textsubscript{FDR} = 0.03), but not Shannon diversity index
(\emph{P}\textsubscript{FDR} = 0.052) across sources of mice (Fig. 2A-B
and Table S3). As expected, clindamycin treatment decreased richness and
Shannon diversity across all sources of mice, and communities started to
recover 1 day post-infection (Fig. 2C-D). Interestingly, significant
differences in diversity metrics across sources
(\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged after both
clindamycin and \emph{C. difficile} infection, with Charles River mice
having higher richness and Shannon Diversity than most of the other
sources (Fig 2C-F and Table S4). While Charles River mice had more
diverse microbiotas, Young and Schloss lab mice were also able to clear
\emph{C. difficile} faster, suggesting microbiota diversity alone does
not explain the observed variation in \emph{C. difficile} colonization
across vendors.

Next, we compared the bacterial communities from the 6 colonies over the
course of the experiment using principal coordinate analysis (PCoA) of
the Theta YC distances. Permutational multivariate analysis of variance
(PERMANOVA) analysis revealed colony source was the major factor
explaining the observed variation across fecal communities
(R\textsuperscript{2} = 0.35, \emph{P} = 0.0001) followed by
interactions between cage and day of the experiment (Movie S1 and Table
S5). Since, the majority of the perturbations happened over the initial
days of the experiment, we decided to focus on the bacterial communities
at baseline (day -1), after clindamycin treatment (day 0), and
post-infection (day 1). For all 3 timepoints, source and the interaction
with cage significantly explained most of the observed community
variation (combined R\textsuperscript{2} = 0.90, 0.99, 0.88,
respectively; \emph{P} = 0.0001; Fig. 3 and Table S6). We also compared
baseline communities across the 2 experiments, and found experiment and
cage significantly explained the observed variation only for the Schloss
and Young lab mouse colonies (Fig. S2A-B and Table S7). However, most of
the vendors also clustered by experiment (Fig. S2C-D, F), suggesting
there was some community variation between the 2 experiments within each
source. Thus, mouse colony source was the factor that explained the most
variation observed in the bacterial communities. Importantly with the
exception of the 2 University colonies, the community of each source
clustered apart from one another suggesting each community had a unique
response to clindamycin treatment and \emph{C. difficile} challenge.

Since there was some variation in microbiota communities between
experiments at baseline, we next looked at how similar the communities
were within the same source and between sources in response to
clindamycin treatment and \emph{C. difficile} challenge (Fig. 4) . The
baseline communities varied most between experiments for Schloss, Young,
and Envigo mice and variation between sources of mice was high (Fig.
4A). Clindamycin treatment reduced variation between experiments within
Schloss, Young and Jackson mice and some of the variation between
sources diminished, particularly for the University of Michigan and
Charles River mice (Fig. 4B). Post-infection, the community variation
started to increase within sources of mice and variation between sources
of mice started to increase towards previous levels (Fig. 4C). By using
mice from multiple sources we were able to increase the number of
microbiota communities we tested with the clindamycin \emph{C.
difficile} colonization mouse model.

After finding differences at the community level, we next identified the
taxa that varied across sources of mice over the initial days of the
experiment. We examined bacterial relative abundances at the operational
taxonomic unit (OTU) and family levels, expecting the number of
differences to be reduced at the family level due to the nature of
bacterial taxonomy (45). Focusing on the baseline communities first,
there were 268 OTUs and 20 families (Tables S8-9) with relative
abundances that varied across colony sources. Clindamycin treatment
reduced the number of taxa with relative abundances that varied across
sources to 18 OTUs and 10 families (Tables S8-9). After \emph{C.
difficile} challenge, there were 44 OTUs and 18 families (Tables S8-9)
with significantly different relative abundances across sources, as the
communities started to recover from antibiotic treatment. In spite of
the experimental perturbations that occurred during these 3 timepoints,
there were 12 OTUs (Fig 5A-C) and 8 families with relative abundances
that consistently varied across colony sources (Fig. 5D-F). Importantly,
some of the taxa that consistently varied across sources also shifted
with clindamycin treatment. For example, \emph{Proteus} increased after
clindamycin treatment, but only in Taconic mice. \emph{Enterococcus} was
primarily found only in mice purchased from commercial vendors and also
increased after clindamycin treatment. In summary, mouse bacterial
communities significantly varied according to colony source throughout
the course of the experiment and a consistent subset of bacterial taxa
remained different across sources regardless of clindamycin and \emph{C.
difficile} challenge.

\textbf{Clindamycin treatment alters a subset of taxa that were found in
all colony sources.} Although there were bacteria that consistently
varied across colony sources, we also wanted to identify the bacteria
that shifted after clindamycin treatment, regardless of colony source.
By analyzing all mice that had sequence data from fecal samples
collected at baseline and after clindamycin treatment, we identified 153
OTUs and 18 families that were altered after clindamycin treatment (Fig.
6 and Tables S10-11). Interestingly, when we compared the list of
significant clindamycin impacted bacteria with the bacteria that
consistently varied across groups over the initial 3 timepoints of our
experiment, we found 3 OTUs (\emph{Lachnospiraceae} (OTU 130),
\emph{Lactobacillus} (OTU 6), \emph{Enterococcus} (OTU 23)) and 3
families (\emph{Porphyromonadaceae}, \emph{Enterococcaceae},
\emph{Lachnospiraceae}) overlapped (Fig. 5, Fig. 6C-D). These findings
demonstrate that clindamycin has a consistent impact on the fecal
bacterial communities of mice from all colony sources and only a subset
of the taxa also varied across colony sources.

\textbf{Source-specific and clindamycin impacted bacteria distinguish
\emph{C. difficile} colonization status in mice.} After identifying taxa
that varied by colony source, changed after clindamycin treatment, or
both, we next wanted to determine which taxa were influencing the
variation in \emph{C. difficile} colonization at day 7 (Fig. 1D, Fig.
S1C). We trained L2-regularized logistic regression models with input
bacterial community data from the baseline, post-clindamycin, and
post-infection timepoints of the experiment to predict \emph{C.
difficile} colonization status on day 7 (Fig. S3A-B). All models were
better at predicting \emph{C. difficile} colonization status on day 7
than random chance (all \emph{P} \(\le\) 5e-15; Table S12). However, the
models trained with OTU level data generally performed better than those
trained with family level data with the expception of the models based
on the post-infection (day 1) communities (Fig. S3C-D). Interestingly,
the model based on the post-clindamycin (day 0) community OTU data
performed significantly better than all other models with an area under
the receiving operator characteristic curve (AUROC) of 0.75
(\emph{P}\textsubscript{FDR} \(\le\) 3.9e-10 for pairwise comparisons;
Table S13). Thus, we were able to use community bacterial relative
abundance data alone to differentiate mice that had cleared \emph{C.
difficile} before day 7 from the mice still colonized with \emph{C.
difficile} at that timepoint. Interestingly, the model built with OTU
relative abundance data post-clindamycin treatment had the best
performance, suggesting how the bacterial community responds to
clindamycin treatment has the greatest influence on subsequent \emph{C.
difficile} colonization dynamics.

Next, to examine the bacteria that were driving each model's
performance, we pulled out the top 20 taxa that had the highest absolute
feature weights in each of the 6 models (Tables S14-15). First, we
looked at OTUs from the model with the best performance that was based
on the post-clindamycin treatment bacterial community data. While most
of the 20 OTUs had low relative abundances on day 0,
\emph{Enterobacteriace}, \emph{Bacteroides} and \emph{Proteus} had high
relative abundances in at least one source of mice and significantly
varied across sources (Fig. 7A). Next, the top 20 taxa from each model
were compared to the list of taxa that varied across colony source at
the same timepoint (Fig. 5 and Tables S8-9) and the taxa that were
altered by clindamycin treatment (Fig. 6 and Table S10-11). We found a
subset of OTUs and families that were important to the model and
overlapped with bacteria that varied by either source, clindamycin
treatment, or both (Fig. S4, S5A-C). Combining the overall results for
the 3 OTU models identified 14 OTUs associated with source, 21 OTUs
associated with clindamycin treatment, and 6 OTUs associated with both
(Fig. 7B). Combining the overall results for the 3 family models
identified 18 families associated with source, 14 families associated
with clindamycin treatment and 8 families associated with both (Fig.
S5D). Several OTUs (\emph{Bacteroides (OTU 2), Enterococcus (OTU 23),
Enterobacteriaceae (OTU 1), Porphyromonadaceae (OTU 7)}) and families
(\emph{Bacteroidaceae, Deferribacteraceae, Enterococcaceae,
Lachnospiraceae, Bifidobacteriaceae, Coriobacteriaceae, Ruminococcaceae,
Verrucomicrobiaceae}) appeared across at least 2 models, so we examined
how the relative abundances of these key taxa varied over the course of
the experiment (Fig. 8 and Fig. S6). Throughout the experiment, there
was at least 1 timepoint where relative abundances of these taxa
significantly varied across sources (Table S16-17). Interestingly, there
were no taxa that emerged as consistently enriched or depleted in mice
that were colonized past 7 days post-infection with \emph{C. difficile}
630, suggesting multiple bacteria influence the time needed to clear the
infection. Together, these results suggest the initial bacterial
communities and their responses to clindamycin influence on the time
needed to clear \emph{C. difficile}.

@@

\textbf{(i) Animals.} All experiments were approved by the University of
Michigan Animal Care and Use Committee (IACUC) under protocol number
PRO00006983. Female C57BL/7 mice were obtained from 6 different colony
sources: The Jackson Laboratory, Charles River Laboratories, Taconic
Biosciences, Envigo, and two colonies at the University of Michigan (the
Schloss lab colony and the Young lab colony). The Young lab colony was
originally established with mice purchased from Jackson, and the Schloss
lab colony was later founded with mice donated from the Young lab. The 4
groups of mice purchased from vendors were allowed to acclimate to the
University of Michigan mouse facility for 13 days prior to starting the
experiment. At least 4 female mice (age 5-10 weeks) were obtained per
source and mice from the same source were primarily housed at a density
of 2 mice per cage. The experiment was repeated once, approximately 3
months after the start of the first experiment.

>>>>

\section{\texorpdfstring{The initial gut microbiota and response to
antibiotic perturbation influence \emph{Clostridioides difficile}
colonization in
mice}{The initial gut microbiota and response to antibiotic perturbation influence Clostridioides difficile colonization in mice}}\label{the-initial-gut-microbiota-and-response-to-antibiotic-perturbation-influence-clostridioides-difficile-colonization-in-mice}

@@

The gut microbiota has a key role in determining susceptibility to
\emph{Clostridioides difficile} infections (CDIs). However, much of the
mechanistic work examining CDIs in mouse models use animals obtained
from a single university colony or vendor. We treated mice from 6
different sources (2 University of Michigan colonies and 4 vendors) with
a single clindamycin dose, followed by a \emph{C. difficile} challenge 1
day later and then measured \emph{C. difficile} colonization levels
through 9 days post-infection. The microbiota was profiled via 16S rRNA
gene sequencing to examine the variation across sources and alterations
due to clindamycin treatment and \emph{C. difficile} challenge. While
all sources of mice were colonized 1-day post-infection, variation
emerged from days 3-7 post-infection with animals from some sources
colonized with \emph{C. difficile} for longer and at higher levels. We
identified bacteria that varied in relative abundance across sources and
throughout the experiment. Some bacteria were consistently impacted by
clindamycin treatment in all sources of mice including
\emph{Lachnospiraceae}, \emph{Ruminococcaceae}, and
\emph{Enterobacteriaceae}. To identify bacteria that were most important
to colonization regardless of the source, we created logistic regression
models that successfully classified mice based on whether they cleared
\emph{C. difficile} by 7 days post-infection using baseline,
post-clindamycin, and post-infection community composition data. With
these models, we identified 4 bacteria that varied across sources
(\emph{Bacteroides}), were altered by clindamycin
(\emph{Porphyromonadaceae}), or both (\emph{Enterobacteriaceae} and
\emph{Enterococcus}). Microbiota variation across sources better
emulates human interindividual variation and can help identify bacterial
drivers of phenotypic variation in the context of CDIs.

@@

\emph{Clostridioides difficile} is a leading nosocomial infection.
Although perturbation to the gut microbiota has been established as a
key risk factor, there is variation in who becomes asymptomatically
colonized, develops an infection, or has an infection with adverse
outcomes. \emph{C. difficile} infection (CDI) mouse models are widely
used to answer a variety of \emph{C. difficile} pathogenesis questions.
However, the inter-individual variation between mice is less than what
is observed in humans, particularly if just one source of mice is used.
In this study, we administered clindamycin to mice from 6 different
breeding colonies and challenged them with \emph{C. difficile}.
Interestingly, only a subset of the bacteria that vary across sources
were associated with how long \emph{C. difficile} was able to colonize.
Future studies examining the interplay between the microbiota and
\emph{C. difficile} should consider using mice from multiple sources to
better reflect human interindividual variation.

@@

Antibiotics are a common risk factor for \emph{Clostridioides difficile}
infections (CDIs), but there is variation in who goes on to develop
severe or recurrent CDIs after exposure (1, 2). Additionally,
asymptomatic colonization, where \emph{C. difficile} is detectable, but
symptoms are absent has been documented in infants and adults (3, 4).
The intestinal microbiome has been implicated in asymptomatic
colonization (5, 6), susceptibility to CDIs (7), and adverse CDI
outcomes (9--12).

Mouse models of CDIs have been a great tool for understanding \emph{C.
difficile} pathogenesis (13). The number of CDI mouse model studies has
grown substantially since Chen et al. published their C57BL/6 model in
2008, which disrupted the gut microbiota with antibiotics to enable
\emph{C. difficile} colonization and symptoms such as diarrhea and
weight loss (14). CDI mouse models have been used to examine
translationally relevant questions regarding \emph{C. difficile},
including the role of the microbiota and efficacy of potential
therapeutics for treating CDIs (15). However, variation in the
microbiome between mice from the same breeding colony is much less than
the variation observed between humans (16, 17). Additionally, studying
the contribution of the microbiota to a particular disease phenotype in
one set of lab mice after the same perturbation could yield a number of
findings of which only a fraction may be driving the phenotype.

In the past, our group has attempted to introduce more microbiome
variation into the CDI mouse model by using a variety of antibiotic
treatments (18--21). An alternative approach to maximize microbiome
variation is to use mice from multiple sources (22, 23). Microbiome
differences between different mouse vendors have been well documented
and shown to influence susceptibility to a variety of diseases (24, 25),
including enteric infections (22, 23, 26--30). Additionally, research
groups have observed different CDI outcomes in mice despite using
similar models and the microbiome has been proposed as one factor
potentially mediating CDI susceptibility and outcomes (13, 18, 21,
31--33). Here we examined how variations in the baseline microbiome and
responses to clindamycin treatment in C57BL/6 mice from six different
sources influenced susceptibility to \emph{C. difficile} colonization
and the time needed to clear the infection.

@@

\textbf{Clindamycin treatment renders all mice susceptible to \emph{C.
difficile} 630 colonization regardless of source.} To test how the
microbiotas of mice from different sources impact colonization dynamics
after clindamycin exposure, we utilized C57BL/6 mice from 6 different
sources: two colonies from the University of Michigan that were split
from each other 10 years ago (the Young and Schloss lab colonies) and
four commercial source: the Jackson Laboratory, Charles River
Laboratories, Taconic Biosciences, and Envigo (which was formerly
Harlan). These 4 vendors were chosen because they represent commonly
used vendors for CDI studies in mice (26, 34--40).

After a 13-day acclimation period for the mice ordered from vendors, all
mice were treated with 10 mg/kg clindamycin via intraperitoneal
injection and one day later challenged with 10\textsuperscript{3}
\emph{C. difficile} 630 spores (Fig. 1A). Clindamycin was chosen because
we have previously demonstrated mice are rendered susceptible, but
consistently cleared the CDI within 9 days (21, 41), clindamycin is
frequently implicated with human CDIs (42), and is also part of the
antibiotic treatment for the frequently cited 2008 CDI mouse model (14).
The day after infection, \emph{C. difficile} was detectable in all mice
at a similar level (median CFU range: 2.2e+07-1.3e+08;
\emph{P}\textsubscript{FDR} = 0.15), indicating clindamycin rendered all
mice susceptible regardless of source (Fig. 1B). Interestingly,
variation in \emph{C. difficile} CFU levels across sources of mice
emerged from days 3-7 post-infection (all \emph{P}\textsubscript{FDR}
\(\le\) 0.019; Fig. 1B and Table S1), suggesting mouse source is
associated with \emph{C. difficile} clearance. We conducted two
experiments approximately 3 months apart and while the colonization
dynamics were similar across most sources of mice, there was some
variation between the 2 experiments, particularly for the Schloss and
Envigo mice (Fig. S1A-B). Although \emph{C. difficile} 630 causes mild
symptoms in mice comparied to other \emph{C. difficile} strains (43), we
also saw that weight change significantly varied across sources of mice
with the most weight lost two days post-infection (Fig. 1C and Table
S2). Importantly, there was also one Jackson and one Envigo mouse that
died between 1- and 3-days post-infection during the second experiment.
Interestingly, mice obtained from Jackson, Taconic, and Envigo tended to
lose more weight (although there was variation between experiments with
Schloss and Envigo mice), have higher \emph{C. difficile} CFU levels and
take longer to clear the infection compared to the other sources of
mice, which was particularly evident 7 days post-infection (Fig. 1B-C,
Fig. S2C-D), when 57\% of the mice were still colonized with \emph{C.
difficile} (Fig. S1E). By 9 days post-infection the majority of the mice
from all sources had cleared \emph{C. difficile} (Fig. S1C) with the
exception of 1 Taconic mouse from the first experiment and 2 Envigo mice
from the second experiment. Thus, clindamycin rendered all mice
susceptible to \emph{C. difficile} 630 colonization, regardless of
source, but there was significant variation in disease phenotype across
the sources of mice.

\textbf{Bacterial communities consistently vary across sources despite
antibiotic and infection perturbations.} Given the well known variation
in mouse microbiomes across breeding colonies (25), we hypothesized that
the variation in \emph{C. difficile} clearance could be explained by
microbiota variation across the 6 sources of mice. We used 16S rRNA gene
sequencing to characterize the fecal bacterial communities from the mice
over the course of the experiment. Since antibiotics and other risk
factors of CDIs are associated with decreased microbiota diversity (44),
we first examined alpha diversity measures across the 6 sources of mice.
Examining the bacterial communities at baseline, prior to clindamycin
treatment, there was a significant difference in the number of observed
OTUs (\emph{P}\textsubscript{FDR} = 0.03), but not Shannon diversity
index (\emph{P}\textsubscript{FDR} = 0.052) across sources of mice (Fig.
2A-B and Table S3). As expected, the clindamycin treatment decreased
richness and Shannon diversity across all sources of mice, and richness
and Shannon diversity started to increase 1 day post-infection (Fig.
2C-D). Interestingly, significant differences in diversity metrics
across sources (\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged
after both clindamycin and \emph{C. difficile} infection, with Charles
River mice having higher richness and Shannon Diversity than most of the
other sources (Fig 2C-F and Table S4). Although the Charles River mice
had more diverse microbiotas, the Young and Schloss lab mice were also
able to clear \emph{C. difficile} faster than the other sources,
suggesting microbiota diversity alone does not explain the observed
variation in \emph{C. difficile} colonization across vendors.

Next, we compared the community structure of mice from the 6 sources
over the course of the experiment using principal coordinate analysis
(PCoA) of the \(\theta_{YC}\) distances. Permutational multivariate
analysis of variance (PERMANOVA) analysis revealed source was the
dominant factor that explained the observed variation across fecal
communities (R\textsuperscript{2} = 0.35, \emph{P} = 0.0001) followed by
interactions between cage (mice from the same sources were housed
together, primarily at a density of 2 mice per cage) and day of the
experiment (Movie S1 and Table S5). Mice that are co-housed have been
shown to have similar gut microbiotas due to coprophagy (45) and since
mice within the same source were housed together, it is not surprising
that cage also contributes to the observed microbiota variation. Since
the majority of the perturbations happened over the initial days of the
experiment, we decided to focus on the communities at baseline (day -1),
after clindamycin treatment (day 0), and post-infection (day 1). For all
3 timepoints, the source and cage interaction significantly explained
most of the observed community variation (combined R\textsuperscript{2}
= 0.90, 0.99, 0.88, respectively; \emph{P} = 0.0001; Fig. 3 and Table
S6). We also compared baseline communities across the 2 experiments, and
found experiment and cage significantly explained the observed variation
only for the Schloss and Young lab mouse colonies (Fig. S2A-B and Table
S7). However, most of the vendors also clustered by experiment (Fig.
S2C-D, F), suggesting there was some community variation between the 2
experiments within each source. Thus, source was the factor that
explained the most variation observed in the bacterial communities.
Importantly with the exception of the Schloss and Young colonies, the
community of each source clustered apart from one another suggesting
each community had a unique response to clindamycin treatment and
\emph{C. difficile} challenge.

Since there was some variation in the microbiota structure of mice from
the same source between experiments at baseline, we next looked at how
similar the communities were within the same source and between sources
in response to clindamycin treatment and \emph{C. difficile} challenge
(Fig. 4) . The baseline communities varied most between experiments for
Schloss, Young, and Envigo mice and variation between sources of mice
was high (Fig. 4A). Clindamycin treatment reduced the variation between
experiments within Schloss, Young and Jackson mice and some of the
variation between sources diminished, particularly for the Schloss,
Young, and Charles River mice (Fig. 4B). Post-infection, the community
variation started to increase within sources of mice and variation
between sources of mice started to increase towards previous levels
(Fig. 4C). By using mice from multiple sources we were able to increase
the number of microbiota communities we tested with the clindamycin
\emph{C. difficile} colonization mouse model.

After finding differences at the community level, we next identified the
bacteria that varied across sources of mice over the initial days of the
experiment. We examined bacterial relative abundances at the operational
taxonomic unit (OTU) level. Focusing on the baseline communities first,
there were 268 OTUs (Table S8) with relative abundances that varied
across sources. Clindamycin treatment reduced the number of taxa with
relative abundances that varied across sources to 18 OTUs (Table S8).
After \emph{C. difficile} challenge, there were 44 OTUs (Table S8) with
significantly different relative abundances across sources, as the
communities started to recover from antibiotic treatment. In spite of
the experimental perturbations that occurred during these 3 timepoints,
there were 12 OTUs (Fig 5A-C) with relative abundances that consistently
varied across sources. Importantly, some of the OTus that consistently
varied across sources also shifted with clindamycin treatment. For
example, \emph{Proteus} increased after clindamycin treatment, but only
in Taconic mice. \emph{Enterococcus} was primarily found only in mice
purchased from commercial vendors and also increased after clindamycin
treatment. In summary, mouse bacterial communities varied significantly
according to source throughout the course of the experiment and a
consistent subset of bacterial taxa remained different across sources
regardless of clindamycin and \emph{C. difficile} challenge.

\textbf{Clindamycin treatment alters a subset of taxa that were found in
all sources.} Although there were bacteria that consistently varied
across sources, we also wanted to identify the bacteria that shifted
after clindamycin treatment, regardless of source. By analyzing all mice
that had sequence data from fecal samples collected at baseline and
after clindamycin treatment, we identified 153 OTUs that were altered
after clindamycin treatment (Fig. 6 and Tables S9). Interestingly, when
we compared the list of significant clindamycin impacted bacteria with
the bacteria that consistently varied across groups over the initial 3
timepoints of our experiment, we found 3 OTUs (\emph{Lachnospiraceae}
(OTU 130), \emph{Lactobacillus} (OTU 6), \emph{Enterococcus} (OTU 23))
overlapped (Fig. 5, Fig. 6C-D). These findings demonstrate that
clindamycin has a consistent impact on the fecal bacterial communities
of mice from all sources and only a subset of the OTUs also varied
across sources.

\textbf{Source-specific and clindamycin impacted bacteria distinguish
\emph{C. difficile} colonization status in mice.} After identifying taxa
that varied by source, changed after clindamycin treatment, or both, we
next wanted to determine which taxa were influencing the variation in
\emph{C. difficile} colonization at day 7 (Fig. 1D, Fig. S1C). We
trained three L2-regularized logistic regression models with input
bacterial community data from the baseline, post-clindamycin, and
post-infection timepoints of the experiment to predict \emph{C.
difficile} colonization status on day 7 (Fig. S3A-B). All models were
better at predicting \emph{C. difficile} colonization status on day 7
than random chance (all \emph{P} \(\le\) 5.2e-31; Table S12).
Interestingly, the model based on the post-clindamycin (day 0) community
OTU data performed significantly better than all other models with an
area under the receiving operator characteristic curve (AUROC) of 0.75
(\emph{P}\textsubscript{FDR} \(\le\) 3.1e-11 for pairwise comparisons;
Table S13). Thus, we were able to use bacterial relative abundance data
alone to differentiate mice that had cleared \emph{C. difficile} before
day 7 from the mice still colonized with \emph{C. difficile} at that
timepoint. Interestingly, the model built with OTU relative abundance
data post-clindamycin treatment had the best performance, suggesting how
the bacterial community responds to clindamycin treatment has the
greatest influence on subsequent \emph{C. difficile} colonization
dynamics.

Next, to examine the bacteria that were driving each model's
performance, we pulled out the top 20 taxa that had the highest absolute
feature weights in each of the 6 models (Tables S14-15). First, we
looked at OTUs from the model with the best performance that was based
on the post-clindamycin treatment bacterial community data. While most
of the 20 OTUs had low relative abundances on day 0,
\emph{Enterobacteriace}, \emph{Bacteroides} and \emph{Proteus} had high
relative abundances in at least one source of mice and significantly
varied across sources (Fig. 7A). Next, the top 20 taxa from each model
were compared to the list of taxa that varied across source at the same
timepoint (Fig. 5 and Tables S8-9) and the taxa that were altered by
clindamycin treatment (Fig. 6 and Table S10-11). We found a subset of
OTUs that were important to the model and overlapped with bacteria that
varied by either source, clindamycin treatment, or both (Fig. S4,
S5A-C). Combining the overall results for the 3 OTU models identified 14
OTUs associated with source, 21 OTUs associated with clindamycin
treatment, and 6 OTUs associated with both (Fig. 7B). Several OTUs
(\emph{Bacteroides (OTU 2), Enterococcus (OTU 23), Enterobacteriaceae
(OTU 1), Porphyromonadaceae (OTU 7)}) appeared across at least 2 models,
so we examined how the relative abundances of these key taxa varied over
the course of the experiment (Fig. 8). Throughout the experiment, there
was at least 1 timepoint where relative abundances of these OTUs
significantly varied across sources (Table S13). Interestingly, there
were no OTUs that emerged as consistently enriched or depleted in mice
that were colonized past 7 days post-infection with \emph{C. difficile}
630, suggesting multiple bacteria influence the time needed to clear the
infection. Together, these results suggest the initial bacterial
communities and their responses to clindamycin influence the clearance
of \emph{C. difficile}.

@@

\textbf{(i) Animals.} All experiments were approved by the University of
Michigan Animal Care and Use Committee (IACUC) under protocol number
PRO00006983. Female C57BL/7 mice were obtained from 6 different sources:
The Jackson Laboratory, Charles River Laboratories, Taconic Biosciences,
Envigo, and two colonies at the University of Michigan (the Schloss lab
colony and the Young lab colony). The Young lab colony was originally
established with mice purchased from Jackson, and the Schloss lab colony
was established 10 years ago with mice donated from the Young lab. The 4
groups of mice purchased from vendors were allowed to acclimate to the
University of Michigan mouse facility for 13 days prior to starting the
experiment. At least 4 female mice (age 5-10 weeks) were obtained per
source and mice from the same source were primarily housed at a density
of 2 mice per cage. The experiment was repeated once, approximately 3
months after the start of the first experiment.
